---
title: Getting started with `ggplot2`
output:
  # xaringan::moon_reader:
  #   nature:
  #     highlightStyle: github
  #     highlightLanguage: r
  #     slideNumberFormat: '%current%/%total%'
  #     ratio: '16:9'
  revealjs::revealjs_presentation:
     transition: slide
     #incremental: true
---

# What is `ggplot2`?

* `ggplot2` is an R-package that was initially developed by Hadley Wickham during his PhD
* `ggplot2` implements a particular interpretation of the "grammar of graphics" by Wilkinson 
* It is one of the most popular packages in R for data visualisation

--- 

# Loading `ggplot2` 

* To use `ggplot2`, you first have to install the package (if you haven't installed it yet)

```{r, eval = FALSE}
install.packages("ggplot2")
```

* Once you have installed it, you can load the package:

```{r}
library(ggplot2)
```
* `ggplot2` is part of the `tidyverse` family so if you load `tidyverse`, you don't need to load `ggplot2` alone

```{r}
library(tidyverse)
```

---

# `ggplot2` basics

```{r plot1, fig.height = 4, fig.width = 4}
ggplot(data = cars, 
       mapping = aes(x = speed, y = dist)) + 
  geom_point()
```

