---
title: Tidy your data
output:
  teachr::teachr_slides:
  revealjs::revealjs_presentation:
     css: [reveal.css, slide.css]
     transition: slide
     keep_md: true
     lib_dir: libs/
     self_contained: false
     #includes: 
     #   before_body: header.html
     #incremental: true
type: slides
chapter: 5
---
```{r setup, include = FALSE}
library(tidyverse)
```

## Tidy (reshape) your data

A tidy data format:

- Each column should hold a single variable
- each row should hold a single observation
- each cell should hold a single value.

Examine `table1`, `table2`, `table3` and `Table5`.

They show the same information in 4 different ways.

---

Table 1

```{r}
table1

```

Note:

Raw data that we acquired can often be messy and complicated. The tidyr package allows you to wrangle messy datasets into nice and tidy ones.

A tidy data format has a rectangular shape which means it has columns, rows, and cells, just like in a spreadsheet. Each column should hold a single variable, each row should hold a single observation and each cell should hold a single value.

The example on the left shows the same information (country, year, population, and number of cases) organised in 3 different ways.
Examine each of the table.

---

Table 2

```{r}
table2

```

Note:

Raw data that we acquired can often be messy and complicated. The tidyr package allows you to wrangle messy datasets into nice and tidy ones.

A tidy data format has a rectangular shape which means it has columns, rows, and cells, just like in a spreadsheet. Each column should hold a single variable, each row should hold a single observation and each cell should hold a single value.

The example on the left shows the same information (country, year, population, and number of cases) organised in 3 different ways.
Examine each of the table.

---

Table 3

```{r}
table3

```

Note:

Raw data that we acquired can often be messy and complicated. The tidyr package allows you to wrangle messy datasets into nice and tidy ones.

A tidy data format has a rectangular shape which means it has columns, rows, and cells, just like in a spreadsheet. Each column should hold a single variable, each row should hold a single observation and each cell should hold a single value.

The example below shows the same information (country, year, population, and number of cases) organised in 3 different ways.
Examine each of the table.

---

Table 5

```{r}
table5

```

Note:

Raw data that we acquired can often be messy and complicated. The tidyr package allows you to wrangle messy datasets into nice and tidy ones.

A tidy data format has a rectangular shape which means it has columns, rows, and cells, just like in a spreadsheet. Each column should hold a single variable, each row should hold a single observation and each cell should hold a single value.

The example below shows the same information (country, year, population, and number of cases) organised in 3 different ways.
Examine each of the table.


---

R functions to tidy the data:

- `separate()`: separating multiple variables in one column to different columns.
- `unite()`: merging multiple columns into one, inserting underscore (`_`) as a separator between the two inputs.
- `separate_rows()`: separating multiple observations in the same cell. We pass the messy column name and a string to use as a separator.

Note:

Sometimes raw data may have multiple variables or multiple values (not the same variable) in the same column, or sometimes you may want to merge multiple columns into one. Here are a few commands that can be helpful in manipulating the data into a tidy format.

---

In table 3, the `rate` column contains both the number of cases and the size of the populations.

We want to separate the `rate` variable into variables `cases` and `population`. 

```{r}

table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)

```

Note:

- The `into` argument expects a vector of column names.
- The `sep` argument expects the non-alphanumeric character (a character that is not a letter or a number) that we split upon. 
- The `convert` argument is set to TRUE here so that R will try and convert to better types after the split.

---

In table 5, the `century` and `year` column can be combined to give a `year_new` column.

```{r}

table5 %>%
  unite(new_year,century, year, sep = "")

```

Note:

The `sep` argument here is what comes in between the values from different columns. If we don't want anything in between, we can use "".

---

In table 6, there are multiple observations in the same cell in each row of `year` and `cases`.

```{r}

table6 <- tibble(country = c("Afghanistan", "Brazil", "China"),
  year = c("1999,2000", "1999,2000", "1999,2000"),
  cases= c("745,2666", "37737,80488", "212258,213766"))

table6
```

---

We would like to separate them.

```{r}
table6 %>% separate_rows(cases, year, sep=',', convert = TRUE)
```

---

## Pivoting

Two common problems in raw data set:

- some of the column names are not really names of variables, but **values** of a variable. 
- when an observation is scattered across multiple rows.

To tackle these problems, you will need `pivot_longer()` and/or `pivot_wider()`

---

**pivot_longer()**

`table4a` is a great example of data that needs `pivot_longer()`. 

Here you see that the case count of each country is separated into two columns because they are from different years.

```{r}
table4a
```

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

Notes: 

- The first argument is the vector of column names that are actually values not variables.
- The second argument is the name of the variable to move the column **names** to.
- The third argument is the name of the variable to move the column **values** to.


---

**pivot_wider()**

In table 2, each observation is a certain country in a certain year.

However, the observation is spread across two rows since there are `cases` and `population`.

```{r}
table2
```


```{r}
table2 %>%
  pivot_wider(names_from = type, values_from = count)
```

Notes:

- The first argument is the column to take variable **names** from, which is `type`.
- The second argument is the column to take variable **values** from, where is `count`.

---

## Missing values

There are various ways to handle missing values (the `NA`s) in your data set.

Tidyr has three functions:

- `fill()`: fill the missing values using the last observation carried forward. It has a `.direction` argument that you can pass `up` or `down`.
- `drop_na()`: drop the rows with observations that are `NA`
- `replace_na()`: replace the `NA` with specific values

---

Let's look at this toy data set. 

We will start with the `fill()` function to impute the `return` column in the `up` direction, then we will impute in the `down` direction.

Can you spot the difference?

```{r}
apple <- tibble(
  year   = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(0.75, 0.39, 0.34,   NA, 0.99, 0.07, 1.88)
)

apple %>%
  fill(return, .direction = "up")

apple %>%
  fill(return, .direction = "down")
```

---

Alternatively, we can try to drop the `NA`s in the return in `apple`.

```{r}
apple %>%
  drop_na(return)
```
 
--- 
 
In some cases, it may make sense if we replace the `NA`s with a certain value. 

For example, suppose we found out that all missing values are actually 0 and not recorded correctly

```{r}
apple %>%
  replace_na(list(return=0))
```
