---
title: Matrices
output:
  revealjs::revealjs_presentation:
     css: [reveal.css, slide.css]
     transition: slide
     keep_md: true
     lib_dir: libs/
     self_contained: false
     #includes: 
     #   before_body: header.html
     #incremental: true
---

```{r setup, include = FALSE}
```
## Matrices

A **matrix** is a rectangular array of numbers written between
rectangular brackets (or large parentheses). For example, matrix `A` is
a $5 \times 3$ matrix (pronounced as `5 by 3`, indicating there are 5
rows and 3 columns in the matrix) is defined as follow:

$$ A = \begin{bmatrix}4&3&2\\7&4&2\\6&3&4\\3&0&0\\2&-2&-3 \end{bmatrix} $$

In R, a matrix is a collection of elements of the same data type
(numeric, character, or logical) arranged into a fixed number of rows
and columns. You can construct a matrix in R using the `matrix()`
function. You can express the matrix `A` above with the following R
code:

`A <- matrix(c(4,3,2,7,4,2,6,3,4,3,0,0,2,-2,-3), byrow=TRUE, nrow=5)`

-   Here the first argument is the collection of elements that R will
    arrange into the rows and columns of the matrix.
-   The second argument `byrow` indicates the arrangement of elements is
    done by filling the matrix by the rows. If you would like to fill
    the matrix by columns, you can use `byrow=FALSE`.
-   The third argument `nrow` indicates that the number of rows in the
    matrix.
    
---

### Matrix Operations

Similar to what you have learned with vectors, the standard operators like `+`, `-`, `/`, `*`, etc. work in an element-wise way on matrices in R. For example, `2 * A` multiplies each element of `A` by two.

```
A <- matrix(c(4,3,2,7,4,2,6,3,4,3,0,0,2,-2,-3), byrow=TRUE, nrow=5)
2*A

```

Consider another matrix `B` defined as `B <- matrix(c(1,1,1,2,2,2,0,0,0,-1,-1,-1,0,0,0), byrow=TRUE, nrow=5)`

`A*B` creates a matrix where each element is the product of the corresponding elements in matrix `A` and matrix `B`.

Those who are familiar with matrices should note that this is **not** the standard matrix multiplication for which you should use `%*%` in R.
 
---
 
### Add rows or columns to a matrix

Sometimes we may want to add new data or information to a matrix with more rows or more columns.
You can add a column or multiple columns to a matrix with the `cbind()` function, which merges matrices and/or vectors together by column. For example, `cbind(A,B)` yields a bigger matrix by merging matrix A and B. The resulting matrix has 5 rows and 6 columns.

```{r}
A <- matrix(c(4,3,2,7,4,2,6,3,4,3,0,0,2,-2,-3), byrow=TRUE, nrow=5)
B <- matrix(c(1,1,1,2,2,2,0,0,0,-1,-1,-1,0,0,0), byrow=TRUE, nrow=5)
big_matrix <- cbind(A, B)
big_matrix

```

Similarly, you can add a row or multiple rows to a matrix with the `rbind()` function. For example, `big_matrix2 <- rbind(A,B)` returns a bigger matrix by putting matrix A on top of matrix B. The resulting matrix has 10 rows and 3 columns.

```{r}
big_matrix2 <- rbind(A,B)
big_matrix2

```

---

### Indexing

Similar to vectors, you can use the square brackets `[ ]` to select one or multiple elements from a matrix. Whereas vectors have one dimension, matrices have two dimensions. You should therefore use a comma to separate the rows you want to select from the columns. For example, 
`A[1:4,3:5]` returns a matrix with the data on the rows 1, 2, 3, 4 and columns 3, 4, 5.

If you want to select all elements of a row or a column, you can do the following:

- `A[,1]` selects all elements of the first column of matrix A.
- `A[1,]` selects all elements of the first row of matrix A.

