---
title: Getting started with `ggplot2`
format: revealjs
---

## What is `ggplot2`?

```{r setup}
knitr::opts_chunk$set(fig.path = "images/",
                      fig.align = "center",
                      fig.width = 4, 
                      fig.height = 4,
                      echo = TRUE)
```

* `ggplot2` is an R-package that was initially developed by Hadley Wickham as part of his PhD
* `ggplot2` implements a particular interpretation of the "The Grammar of Graphics" by Leland Wilkinson 
* It is one of the most popular packages in R for data visualisation and widely used in scientific outputs, reports, and even news articles

---

## Installing and loading `ggplot2` 


* To use `ggplot2`, you first have to install the package 

```{r, eval = FALSE}
install.packages("ggplot2")
```

* Once you have installed it, you can load the package:

```{r}
library(ggplot2)
```
* `ggplot2` is part of the `tidyverse` family so if you load `tidyverse`, you don't need to load `ggplot2` like above

```{r}
library(tidyverse)
```

---

## Motivating data 1 💎 diamonds

* Let's have a look at the `diamonds` data which contains information about different attributes of diamonds

```{r}
data(diamonds, package = "ggplot2")
diamonds
```


---

## Initialising the plot

```{r plot1-initial}
ggplot(data = diamonds) 
```

---

## Mapping data variables to aesthestics

```{r plot1-map}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) 
```

---

## Adding a layer

* Scatter plot of price vs carat of diamonds
* Each point correponds to a dimaond

```{r plot1-layer}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_point()
```

---

## Adding a different layer

* A hexagonal heatmap where the hexagon shows the count of observations within the region

```{r plot1-boxplot}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_hex()
```

---



