---
title: Multiple subplots in `ggplot2`
format: 
  revealjs:
    #self-contained: true
    logo: images/monash-stacked-blue-rgb-transparent.png
    slide-number: true
    show-slide-number: all
    controls: true
    width: 1280
    height: 720
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(#fig.path = "images/chapter6-06/",
                      fig.align = "center",
                      fig.width = 10, 
                      fig.height = 4,
                      cache.path = "cache/",
                      cache = FALSE,
                      echo = TRUE)
```


## Facets 

* Faceting in `ggplot2` refers to partitioning of the plot region to multiple panels each showing a different subset of the data.
* There are three types of faceting:

  * `facet_null()` (the default),
  * `facet_wrap()`, and
  * `facet_grid()`.
  
Notes: 

* We don't really use `facet_null()` much.

---

## Illustrative data 🌱 growth of soybean

```{r}
data(davidian.soybean, package = "agridat")
str(davidian.soybean)
```

Notes:

* The data contains:
  * `plot`: the plot code
  * `variety`: the code (`F` = commercial and `P` = experimental) for which soybean variety
  * `year`: the year it was grown (1988-1990)
  * `day` days after planting
  * `weight` weight of soybean leaves

---

## `facet_wrap()`

* `facet_wrap()` divides the data into mutually exclusive subset based on the input variable and create subplots based on the data in the corresponding panel.

```{r facet-wrap-demo, fig.height = 2}
library(ggplot2)
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point() +
  facet_wrap(~year)
```

---

## Wrapping many panels

* `facet_wrap` "wraps" the panels into 2d if you have many panels

```{r facet-wrap-demo-wrap}
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point() +
  facet_wrap(~plot)
```

---

## Wrapping by a continuous variable

* Discretize  continuous variable using `cut()`, `cut_width()`, `cut_interval()` or `cut_number()`.

```{r facet-wrap-demo-cont, fig.height = 3}
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point() +
  facet_wrap(~cut_number(day, 4))
```

Notes: 

* The cut functions categorize the continuous variable into discrete levels.

---

## Changing the scales for each panel

* By default the scales are shared across the panels but you can choose to have scales built independently for each panel:

```{r facet-wrap-demo-cont-scales, fig.height = 3}
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point() +
  facet_wrap(~cut_number(day, 4), scales = "free")
```


Note:

* If you want the scale to be independent only in one dimension then you can use `scales = "free_x"` and `scales = "free_y"`.

---

## Wrapping with more than one variables

* If you want to wrap the facet by more than one variable, you can combine the variables by using `+`:

```{r facet-wrap-demo-interaction, fig.height = 3}
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point() +
  facet_wrap(~ year + variety)
```


Notes:

* Alternatively, you can use `vars()` instead of `+`.

---

## `facet_grid()`

* Alternatively, you can use `facet_grid()` to lay out the panels in a 2d grid:

```{r facet-grid-demo, fig.height = 3}
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point() +
  facet_grid(variety ~ year)
```

---

## Missing facet variables

* If a layer data is missing the faceting variable(s), then the layer will be shown across all panels based on all of the layer data

```{r facet-all, fig.height = 2.5}
ggplot(davidian.soybean, aes(day, weight)) + 
  geom_point(data = dplyr::select(davidian.soybean, -c(variety, year)),
             color = "grey70") +
  geom_point() +
  facet_grid(variety ~ year)
```

Notes:

* This is a nice way to creating a "shadow" across all panels.

---

## Combining plots with `patchwork`

* We use the `patchwork` package for combining plots.

```{r patchwork}
library(patchwork)
```
* Combining multiple `ggplot` objects is then just a matter of using `+` (side-by-side) or `/` (top-to-bottom).


---


## Demo with `patchwork`

```{r patchwork-demo, fig.height = 3}
g1 <- ggplot(davidian.soybean, aes(day, weight)) +
  geom_point(aes(color = as.factor(year))) 
g2 <- ggplot(davidian.soybean, aes(as.factor(year), weight)) +
  geom_jitter(aes(color = as.factor(year))) 
g1 + g2
```

---

### Combining plot layouts

* `patchwork` can collect the same guides across the plots as well as automatically add tag labels

```{r patchwork-layouts}
g1 + g2 + 
  plot_layout(guides = "collect") + 
  plot_annotation(tag_levels = "A")
```

---

### Plot layout

* You can combine the use of `+` and `/` to layout plots horizontally or vertically using `(` to group rows or columns




```{r patchwork-layout}
(g1 + g2) / (g2 + g1) + 
  plot_layout(guides = "collect")
```

---

## Flexible plot layout

* If you want even more control over the plot layout, then you can specify a string with the structure you want like below:

```{r patchwork-design, eval = FALSE}
design <- "
 11###
 11###
 #2222"

g1 + g2 + 
  plot_layout(guides = "collect",
              design = design)
```

Notes:

* Notice the empty spots in this plot output.
* How does it align the the `design`?

```{r patchwork-design, echo = FALSE}
```


---

## Summary

* You can view multiple subplots for different partitions of the data using `facet_wrap()` or `facet_grid()`.
* You can combine plots using `patchwork` package.
