---
title: Simple linear regression
#execute:
#  keep-md: true
format: 
  #teachr::teachr_slides:
  revealjs:
    #self-contained: true
    logo: images/monash-stacked-blue-rgb-transparent.png
    slide-number: true
    show-slide-number: all
    controls: true
    width: 1280
    height: 720
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(#fig.path = "images/chapter8-01/",
                      fig.align = "center",
                      fig.width = 10, 
                      fig.height = 4,
                      cache.path = "cache/",
                      cache = FALSE,
                      echo = TRUE)
library(tidyverse)
library(broom)
library(grid)
library(patchwork)
```

<style>
.columns {
  display: flex
}
</style>


## Simple linear regression

* Simple linear regression refers to fitting a straight line $y = \beta_0 + \beta_1 x$ to a set of bivariate data.

<div class="columns">

<div class="column" style="width:70%">

```{r slr-plot, eval = FALSE}
ggplot(cars, aes(speed, dist)) +
  geom_point() + 
  geom_abline(intercept = -17, 
              slope = 4, size = 1.5,
              color = "indianred") + 
  geom_abline(intercept = -5, 
              slope = 3, size = 1.5,
              color = "cadetblue")  + 
  geom_abline(intercept = 80, 
              slope = -3.5, size = 1.5,
              color = "seagreen")
```
</div>

<div class="column" style="width:30%">

```{r slr-plot, echo = FALSE, fig.width = 4}
```

</div>

</div>

Notes:

* Three lines are shown in the plot: 
  * $y = -17 + 4x$
  * $y = -5 + 3x$
  * $y = 80 - 3.5x$
* Which of these three lines do you think explains the relationship between $y$ and $x$ the best?



---

## Residuals

* A strategy to find the best line of fit is generally based on residuals.
* In simple linear regression, the **residuals** are the vertical distance of the observations from the fitted line.

```{r residuals, echo = FALSE, fig.width = 12}
fit <- augment(lm(dist ~ speed, cars))
g1 <- ggplot(fit, aes(speed, dist)) + 
  geom_point() + 
  geom_line(aes(y = .fitted), color = "#006dae", size = 1.5) +
  geom_segment(aes(x = speed, xend = speed,
                   y = .fitted, yend = dist), 
               color = "indianred", size = 1.2)

g2 <- ggplot() + theme_void() +
  annotate("text", x = 0.5, y = 0.5, label = "Rotate and\nproject →", size = 10)

g3 <- ggplot(fit, aes(speed, .resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "#006dae", size = 1.5) +
  geom_segment(aes(x = speed, xend = speed,
                   y = .resid, yend = 0), 
               color = "indianred", size = 1.2)

g1 + g2 + g3
```



---

## Least squares

$$y = \beta_0 + \beta_1 x$$

* **Method of least squares** to estimate the parameters, $\beta_0$ (intercept) and $\beta_1$ (slope).
* Method of least squares estimate parameters such that it minimizes the residual sum of the squares (RSS).
* **Residuals** = observed $-$ fitted value.
* **Fitted value** = the predicted response $(y)$ value from the fitted model using predictors $(x)$ from the data

Notes:

* Unless stated otherwise, simple linear regression uses the **method of least squares** to estimate the model parameters.
* Method of least squares estimates the model parameters by finding parameters that minimise the sum of the squares of the residual.
* More formally, residuals are the difference of observed value from the fitted value.
* Fitted values are the predicted response value under the fitted model for the predictors in the data.



---

## Sum of the squares of the residuals

```{r plot-fits, echo = FALSE, fig.width = 12, fig.height = 6}
df <- cars %>% 
  mutate(fit1 = -17 + 4 * speed,
         fit2 = -5 + 3 * speed,
         fit3 = 80 - 3.5 * speed)

rss_plot <- function(fit, label) {
  df %>% 
    mutate(res = dist - df[[fit]],
           ind = res > 0) %>% 
    ggplot(aes(speed, dist)) +
    geom_rect(aes(xmin = ind * (speed - abs(res)) + 
                         (1 - ind) * (speed + abs(res)), 
                  xmax = speed,
                  ymin = df[[fit]], ymax = dist), fill = "indianred",
                 color = "indianred", size = 1.2, alpha = 0.2) +
    geom_line(aes(y = df[[fit]]), color = "#006dae", size = 1.5) +
    geom_point() + 
    annotate("text", label = label, x = 10, y = 70, size = 6, color = "#006dae", fontface = "bold") +
    ggtitle(paste0("RSS = ", scales::comma(sum((df[[fit]] - df$dist)^2)))) +
    coord_cartesian(xlim = c(min(df$speed), max(df$speed)),
                    ylim = c(min(df$fit3), max(df$fit1)))
}

g1 <- rss_plot("fit1", "y = -17 + 4x")
g2 <- rss_plot("fit2", "y = -5 + 3x")
g3 <- rss_plot("fit3", "y = 80 - 3.5x")

g1 + g2 + g3
```

Notes:

* The sum of the red squares are the residual sum of squares (RSS) -- the actual RSS can be seen on top of the plot
* It doesn't look like squares because the scale on the x- and y-axis differ
* There is a closed-form solution to finding parameters that minimise the residual sum of the squares. 
* You can use calculus to find this or consult any standard first year statistics text book. 
* We expect you to know most of these results and focus on fitting linear models in R now.



---

## Linear regression in R


* In R, you can fit a linear regression using the method of least squares using the `lm()` function.
* The input for `lm` is a **two-sided formula** that symbolically represents the linear model. 
* For example, `y ~ 1 + x` corresponds to a linear model 

$$\color{#006dae}{y_i} = \beta_0 \cdot \color{#006dae}{1} + \beta_1 \cdot \color{#006dae}{x_i} + e_i, \qquad\text{for } i = 1, ..., n, \text{ and assuming }e_i \sim N(0, \sigma^2).$$

Notes:

* The intercept term `1` is included by default.
* So `lm(y ~ 1 + x)` is the same as `lm(y ~ x)` . 

---

## Fitting the linear regression with R

* To fit a linear regression, it's best to parse the `data.frame` into the `data` argument.
* The variables in the formula refers to the variables in the input data.

```{r}
fit <- lm(dist ~  speed, data = cars)
```

* We'll denote the model like below:

$$\texttt{dist} = \beta_0 + \beta_1 \texttt{speed}.$$

Notes:

$$\texttt{dist} = \beta_0 + \beta_1 \texttt{speed}.$$

* Above is not a mathematically rigorous representation . 
* If you're representing the model mathematically, you should denote the model like below

$$y_i = \beta_0 + \beta_1 x_i + e_i$$
for $i = 1, ..., 50$ where: 

* $y_i$ is the stopping distance of the $i$-th car,
* $x_i$ is the speed (in miles per gallon) of $i$-th car,
* $\beta_0$ and $\beta_1$ are the intercept and slope, and 
* we assume the error $e_i \sim NID(0, \sigma^2)$. 

---

## Getting the model parameter estimates

* To get the leqast squares estimates, referred to also as **coefficients**, you can use `coef` function:

```{r}
coef(fit)
```

* You can get an estimate of $\sigma$ by:

```{r}
sigma(fit)
```

---

## Extracting residual

```{r}
residuals(fit)
```


---

## Model summaries 

```{r}
summary(fit)
```

---

## Exracting model summary data

* You can see useful information from  `summary(fit)` but it is not easy to extract it as a tidy data. 

* The package `broom` has handy functions to get the summary values out in a tidy format:

```{r}
broom::tidy(fit)
```
```{r}
broom::glance(fit)
```

---

## Augment data with model information

* You can also easily augment the data with model information, e.g. fitted values and residuals.

```{r}
broom::augment(fit)
```

---

## Transformations

* You can apply transformations directly in the model formula:


```{r}
fit_transform <- lm(log10(dist) ~  sqrt(speed), data = cars)
```

$$\sqrt{\texttt{dist}} = \beta_0 + \beta_1 \log_{10}(\texttt{speed}).$$

* Whether this model is sensible is another story.

```{r}
coef(fit_transform)
```


---

## Predictions from the fitted model


$$\widehat{\texttt{dist}} = -17.58 + 3.93\cdot \texttt{speed}$$


* The `predict` function can return predicted response values from a fitted model:

```{r}
predict(fit, data.frame(speed = c(5, 10)))
```

Notes:

```{r predict-plot, echo = FALSE, fig.width = 5}
pred <- data.frame(speed = c(5, 10), dist = predict(fit, data.frame(speed = c(5, 10))))
ggplot(cars, aes(speed, dist)) +
  geom_smooth(method = lm, se = FALSE, color = "#006dae", size = 1.5, formula = y ~ x) +
  geom_point() +
  geom_point(data = pred, color = "indianred", size = 3, shape = 10) +
  scale_y_continuous(breaks = seq(0, 120, by = 20)) +
  geom_segment(data = data.frame(x = c(5, -Inf, 10, -Inf),
                                 xend = c(5, 5, 10, 10),
                                 y = c(-Inf, pred$dist[1], -Inf, pred$dist[2]),
                                 yend = c(pred$dist[1], pred$dist[1], pred$dist[2], pred$dist[2])), 
               aes(x = x, xend = xend, y = y, yend = yend),
               color = "indianred", linetype = "dashed")
  
```

---

## Predictions from the model with transformations


$$\log_{10}(\widehat{\texttt{dist}}) = -4.42 + 9.23\cdot \sqrt{\texttt{speed}}$$


* No need to transform the covariates with `predict`:

```{r}
predict(fit_transform, data.frame(speed = c(5, 10)))
```

* But the prediction is the response $\log_{10}({\widehat{\texttt{dist}}})$, so you need to apply the inverse transformation to get the value of $\widehat{\texttt{dist}}$.




Notes:

* Prediction of $\texttt{dist}$:

```{r}
10^predict(fit_transform, data.frame(speed = c(5, 10)))
```


```{r predict-plot-transform, echo = FALSE, fig.width = 5, warning = FALSE, message = FALSE}
pred <- data.frame(speed = c(5, 10), dist = 10^predict(fit_transform, data.frame(speed = c(5, 10))))
ggplot(cars, aes(speed, dist)) +
  geom_smooth(method = lm, se = FALSE, color = "#006dae", size = 1.5, formula = y ~ sqrt(x)) +
  geom_point() +
  geom_point(data = pred, color = "indianred", size = 3, shape = 10) +
  scale_y_log10() +
  geom_segment(data = data.frame(x = c(5, -Inf, 10, -Inf),
                                 xend = c(5, 5, 10, 10),
                                 y = c(1, pred$dist[1], 1, pred$dist[2]),
                                 yend = c(pred$dist[1], pred$dist[1], pred$dist[2], pred$dist[2])), 
               aes(x = x, xend = xend, y = y, yend = yend),
               color = "indianred", linetype = "dashed") +
  coord_trans(y = scales::exp_trans(10))
  
```

---

## Summary 

* You've seen how the `lm` function is used to fit linear models.
* You can use functions like `summary`, `coef`, `sigma`, `broom::tidy`, `broom::glance`, and `broom::augment` to get information about the fitted model.
* Transformations can be applied directly in model formula.
* The `predict` function is useful get the prediction of the response for a new set of data. 
