---
title: R Chunk Settings  
output:
  teachr::teachr_slides
type: slides
chapter: 8
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = "images/chapter9/",
                      fig.align = "center",
                      fig.width = 2, 
                      fig.height = 2,
                      cache.path = "cache/",
                      cache = FALSE,
                      echo = TRUE)

library(icons)
library(tidyverse)
```

### Introduction to R Chunk 

````markdown
`r ''````{r, this-is-an-r-chunk}
```
````

Here is an example for R chunk. An R chunk starts with three ` ```{r} ` and ends with ` ``` `. After ` r,` is where you can name the code chunk. Notice that the code chunk name must not include breaks.


---

### `include` 

````markdown
`r ''````{r, this-is-an-r-chunk,include = TRUE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```
````

```{r, includeture,include=TRUE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```

Note: When you set `include = TRUE`, it will print out both code and results. Alternatively, if you wish not to include the code and results from this R chunk, set `include = FALSE`. R Markdown still runs the code and you can use the results from this code chunk in other chunks. 

---

### `echo`

````markdown
`r ''````{r, this-is-an-r-chunk,echo = FALSE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```
````


```{r, echofalse, echo = FALSE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```


Note: When you set `echo = FALSE`, it will only print out the results from this code chunk but prevents code from appearing in the finished file. This is a useful way to embed figures.

--- 

### `eval` 


If you only want to print out the code but not run the code, you can set `eval = FALSE`. This is helpful when you are  

---

### `message` and ` warning`


````markdown
`r ''````{r, this-is-an-r-chunk,messge = FALSE, warning = FALSE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```
````


Note: When you are generating an assignment or report, you would not want to include messages and warning to be printed out in the final outputs. Set both `message = FALSE` and `warning = FALSE` will prevent messages and warnings to be printed out. 


--- 

### 


