---
title: Multiple linear regression
#execute:
#  keep-md: true
format: 
  #teachr::teachr_slides:
  revealjs:
    #self-contained: true
    logo: images/monash-stacked-blue-rgb-transparent.png
    slide-number: true
    show-slide-number: all
    controls: true
    width: 1280
    height: 720
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center",
                      fig.width = 10, 
                      fig.height = 4,
                      cache.path = "cache/",
                      cache = FALSE,
                      echo = TRUE)
library(tidyverse)
```


## Assumptions for linear models

For $i = 1, ..., n$, 


$$y_i = \beta_0 + \beta_1 x_{i1} + ... +  \beta_kx_{ik} + e_i$$
where we assume <span style="color:red">$e_i$ are i.i.d. $NID(0, \sigma^2)$</span>.

Notes:


* i.i.d. =  independent and identically distributed 




---

## Illustrative data: marketing 

```{r}
data(marketing, package = "datarium")
str(marketing)
```


Notes: 

* The data contains the impact of three advertising medias on sales.
* The variables in the data are:
  * `youtube`: the amount of advertising budget spent on youtube advertising (in $1000),
  * `facebook`: the amount of advertising budget spent on facebook advertising (in $1000),
  * `newspaper`: the amount of advertising budget spent on newspaper advertising (in $1000), and 
  * `sales`: the amount of sales in ($1000).
* Each observation corresponds to one particular trial period. 
* There are 200 trial periods in total.

---

## Model diagnostics for linear models

1. Plot scatter plot between pairs of continuous variables:

```{r ggpairs, message = FALSE, warning = FALSE}
GGally::ggpairs(marketing)
```

