---
title: Getting started with `ggplot2`
execute:
  keep-md: true
format: 
  revealjs:
    logo: images/monash-stacked-blue-rgb-transparent.png
    slide-number: true
    show-slide-number: all
    # below doesn't seem to be working
    #width: wide
type: slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = "images/chapter7-02/",
                      fig.align = "center",
                      fig.width = 4, 
                      fig.height = 4,
                      cache.path = "cache/",
                      cache = TRUE,
                      echo = TRUE)
```

## What is `ggplot2`?



* `ggplot2` is an R-package that was initially developed by Hadley Wickham as part of his PhD
* `ggplot2` implements a particular interpretation of the "The Grammar of Graphics" by Leland Wilkinson 
* It is one of the most popular packages in R for data visualisation and widely used in scientific outputs, reports, and even news articles

---

## Installing and loading `ggplot2` 


* To use `ggplot2`, you first have to install the package 

```{r, eval = FALSE}
install.packages("ggplot2")
```

* Once you have installed it, you can load the package:

```{r}
library(ggplot2)
```
* `ggplot2` is part of the `tidyverse` family so if you load `tidyverse`, you don't need to load `ggplot2` like above

```{r}
library(tidyverse)
```

---

## Motivating data 1 💎 diamonds

* Let's have a look at the `diamonds` data which contains information about different attributes of diamonds

```{r}
data(diamonds, package = "ggplot2")
diamonds
```


---

## Initialising the plot


```{r plot-initial}
ggplot(data = diamonds) 
```

* When there is no layer, it produces a blank layer like top plot.

---

## Mapping data variables to aesthestics

```{r plot-mapping}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) 
```

* This get the scale ready but no layer is defined so nothing is rendered in the panel.

---

# `geom` layers



---

## A scatterplot with `geom_point()`

```{r geom-point}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_point()
```

* Scatter plot of price vs carat of diamonds
* Each point correponds to a dimaond

---

## A hexagonal 2D heatmap with `geom_hex()`

```{r geom-hex}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_hex()
```

* The hexagon shows the count of observations within the region.

---

## A 2D heatmap with `geom_bin_2d()`

```{r geom-bin2d}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_bin_2d()
```

* Similar to `geom_hex()` but the shapes are boxes.

---

## Count of overlapping points with `geom_count()`

```{r geom-count}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_count()
```


---

## A boxplot with `geom_boxplot()`

```{r geom-boxplot}
ggplot(data = diamonds, 
       mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```


---

## A violin plot with `geom_violin()`

```{r geom-violin}
ggplot(data = diamonds, 
       mapping = aes(x = cut, y = price)) +
  geom_violin()
```

---

## A dot plot with `geom_dotplot()`

```{r geom-dotplot}
ggplot(data = mtcars, 
       mapping = aes(x = mpg)) +
  geom_dotplot()
```

---

## A histogram with `geom_histogram()`

```{r geom-histogram}
ggplot(data = mtcars, 
       mapping = aes(x = mpg)) +
  geom_histogram()
```

---

## A frequency polygon with `geom_freqpoly()`

```{r geom-freqpoly}
ggplot(data = mtcars, 
       mapping = aes(x = mpg)) +
  geom_freqpoly()
```



---

## A 2D contour plot with `geom_contour()`

```{r}
ggplot(data = faithfuld, 
       mapping = aes(x = waiting, y = eruptions, z = density)) + 
  geom_contour()
```

---

## A 2D contour plot with `geom_contour_filled()`

```{r}
ggplot(data = faithfuld, 
       mapping = aes(x = waiting, y = eruptions, z = density)) + 
  geom_contour_filled()
```

---

## A 2D contour plot with `geom_density_2d()`

```{r}
ggplot(data = faithful, 
       mapping = aes(x = eruptions, y = waiting)) + 
  geom_density_2d()
```

---

## A 2D contour plot with `geom_density_2d_filled()`

```{r}
ggplot(faithful, aes(x = eruptions, y = waiting)) +
 geom_density_2d_filled()
```

---

## Density plot with `geom_density()`

```{r}
ggplot(data = diamonds, 
       mapping = aes(x = carat)) +
  geom_density()
```

---

## `geom_line`

```{r}
ggplot(economics, aes(date, unemploy)) + 
  geom_line()
```

---

```{r}
ggplot(economics, aes(unemploy/pop, psavert)) +
  geom_path()
```

---

```{r}
ggplot(economics, aes(unemploy/pop, psavert)) +
  geom_step()
```


---

```{r}
world <- map_data("world")
ggplot(world, aes(long, lat, group = group)) +
  geom_polygon()
```

---

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```


---

```{r}
ggplot(mtcars, aes(wt, mpg, label = rownames(mtcars))) +
  geom_text()
```

---

```{r}
ggplot(mtcars, aes(wt, mpg, label = rownames(mtcars))) +
  geom_label()
```

---

```{r}
ggplot(faithfuld, aes(waiting, eruptions)) +
 geom_raster(aes(fill = density))
```

---

```{r}
df <- data.frame(
  x = rep(c(2, 5, 7, 9, 12), 2),
  y = rep(c(1, 2), each = 5),
  z = factor(rep(1:5, each = 2)),
  w = rep(diff(c(0, 4, 6, 8, 10, 14)), 2)
)
ggplot(df, aes(x, y)) +
  geom_tile(aes(fill = z), colour = "grey50")
ggplot(df, aes(x, y, width = w)) +
  geom_tile(aes(fill = z), colour = "grey50")
ggplot(df, aes(xmin = x - w / 2, xmax = x + w / 2, ymin = y, ymax = y + 1)) +
  geom_rect(aes(fill = z), colour = "grey50")
```

