---
title: Getting started with `ggplot2`
output:
  # xaringan::moon_reader:
  #   nature:
  #     highlightStyle: github
  #     highlightLanguage: r
  #     slideNumberFormat: '%current%/%total%'
  #     ratio: '16:9'
  revealjs::revealjs_presentation:
     css: [reveal.css, slide.css]
     transition: slide
     keep_md: true
     lib_dir: libs/
     self_contained: false
     #includes: 
     #   before_body: header.html
     #incremental: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = "images/")
```

# What is `ggplot2`?

* `ggplot2` is an R-package that was initially developed by Hadley Wickham during his PhD
* `ggplot2` implements a particular interpretation of the "grammar of graphics" by Wilkinson 
* It is one of the most popular packages in R for data visualisation

--- 

# Loading `ggplot2` 

```r
library(ggplot2)
```

* To use `ggplot2`, you first have to install the package (if you haven't installed it yet)

```{r, eval = FALSE}
install.packages("ggplot2")
```

```{r}
2 + 3
```


* Once you have installed it, you can load the package:

```{r}
library(ggplot2)
```
* `ggplot2` is part of the `tidyverse` family so if you load `tidyverse`, you don't need to load `ggplot2` alone

```{r}
library(tidyverse)
```

---

# `ggplot2` basics

```{r plot1, fig.height = 4, fig.width = 4}
ggplot(data = cars, 
       mapping = aes(x = speed, y = dist)) + 
  geom_point()
```






---

## Average weekday for a full-time university student

```{r}
data <- data.frame(activity = c("Education", "Work", "Sleep", "Leisure", "Other"),
                   hours = c(3.5, 2.3, 8.8, 4, 5.4))

data
```

---

## Plotting with R (base version)


```{r base-barplot, fig.width = 7}
barplot(hours ~ activity,
        data = data)
```


---

## Plotting with R (ggplot2 version)


```{r ggplot-barplot, fig.width = 7, cache = FALSE}
library(ggplot2)
ggplot(data = data, 
       mapping = aes(x = activity, 
                     y = hours)) +
  geom_col()
```


