---
title: Layers in `ggplot2`
format: revealjs
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = "images/",
                      fig.align = "center",
                      fig.width = 4, 
                      fig.height = 4,
                      echo = TRUE)
library(ggplot2)
```

## Layer

* A layer is usually added using functions with prefix `geom_` and `stat_`
* A layer is comprised of:
   - `data` 
   - `mapping` 
   - statistical tranformation (`stat`)
   - geometric object (`geom`)
   - position adjustment (`position`)
* `data` and `mapping` are often inherited from `ggplot`



## The boxplot layer

```{r plot1-cut}
ggplot(data = diamonds,
       mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

---

## You can add more than one layer

```{r plot1-layer2}
ggplot(data = diamonds,
       mapping = aes(x = cut, y = price)) +
  geom_violin() + 
  geom_boxplot(width = 0.1)
```

---

## The order of the layer matters

```{r plot1-layer-order}
ggplot(data = diamonds,
       mapping = aes(x = cut, y = price)) +
  geom_boxplot(width = 0.1) +
  geom_violin()
```

---

## Inheriting data and mapping

* A layer inherits the data and mapping from the initialised ggplot object by default 
* You can add layer specific data and mapping


```{r plot2-inheritance, fig.width = 8}
ggplot(data = diamonds) +
  geom_violin(aes(x = cut, y = price, color = cut)) + 
  geom_boxplot(aes(x = cut, y = price),
               width = 0.1)
```

