---
title: 'Chapter 7: Data Visualisation with R'
description:
  'This chapter will give you an introduction to the R graphics system and teach you how to get started with using the `ggplot2` package for drawing all kind of plots.'
prev: /chapter6
next: /chapter8
type: chapter
id: 7
output:
  html_document:
    self_contained: false
  #md_document
---

```{r setup, include = FALSE}
library(teachr)
knitr::opts_chunk$set(fig.path = "../static/ch7/",
                      fig.retina = 2,
                      cache = FALSE,
                      cache.path = "cache/")
# `slides` and `opts` (now called `mc_opts` functions now live in the `teachr` package
# install latest at https://github.com/numbats/teachr

```


<exercise id="1" title="R Graphics" type="slides">

`r slides("chapter7_01_rgraphics")`

</exercise>

<exercise id="2" title="Test your knowledge of R Graphics">

Which of the following is a contributed R package?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("ggplot2" = "That's correct! ggplot2 was developed by Hadley Wickham as part of his PhD.",
         "grid" = "No, this is a core package so it's already installed.",
         "graphics" = "No, this a core package and loads automatically when you launch R.",
         "grDevices" = "No, this a core package and loads automatically when you launch R.",
         correct = 1))
```

Which R package actually renders the graphics in R?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("ggplot2" = "No, that's incorrect.",
         "grid" = "No, that's incorrect.",
         "graphics" = "No, that's incorrect.",
         "grDevices" = "Yes, that's right!",
         correct = 4))
```

Remember that there are two primary graphic models in R: the base and grid graphics. Which one is `ggplot2` using?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("base" = "No, that's incorrect.",
         "grid" = "Yes that's right! Well done!",
         correct = 2))
```


</exercise>



<exercise id="3" title="Getting started with `ggplot2`" type="slides">

`r slides("chapter7_02_ggplot2intro")`

</exercise>


<exercise id="4" title="Make basic plots with `ggplot2`">

```{teachr 07_04, outdir = "../exercises"}
library(ggplot2)
```


<codeblock id="07_04">

This is a hint.

</codeblock>


</exercise>


<exercise id="5" title="Multiple layers in `ggplot2`">
`r slides("chapter7_03_ggplot2layers")`
</exercise>


<exercise id="7" title="Scales and guides in `ggplot2`">
`r slides("chapter7_04_ggplot2scales")`
</exercise>

<exercise id="9" title="Multiple plots in one figure">
`r slides("chapter7_05_ggplot2facet")`
</exercise>

<exercise id="11" title="Customise the look with themes in `ggplot2`">
`r slides("chapter7_06_ggplot2theme")`
</exercise>
