---
title: 'Chapter 6: Data Wrangling'
description:
  'In this chapter, you will learn to use tidyr and dplyr from the tidyverse in R.'
prev: /chapter5
next: /chapter7
type: chapter
id: 6
output:
  teachr::teachr_chapter:
  html_document:
    self_contained: false
    keep_md: true
---

```{r setup, include = FALSE}
library(teachr)
library(testthat)
```


<exercise id="1" title="Introduction to tidyverse">

## Introduction to tidyverse

The tidyverse is a collection of R packages that transform and visualize data. All packages of the tidyverse share very similar syntax and ideologies.

Tidyverse includes the following packages:

- `dplyr`: grammar of data manipulation
- `tidyr`: reshape your data
- `ggplot2`: grammar of graphics
- `readr`: read rectangular data
- `tibble`: re-imagining of data frame
- `stringr`: working with string data

and many more...


There are many other packages with more specialized functions in the tidyverse. In this chapter, we focuses on two packages: **tidyr** and **dplyr**.

**dplyr** is used for grammar of data manipulation. **tidyr** helps you to create tidy data or data where each variable is in a column, each observation is a row and each value is a cell. You can use them to solve the most common data manipulation challenges.

You can install the core tidyverse packages with:
`install.packages("tidyverse")`

Then, to load the package:
```{r}
library(tidyverse)
```

You can also load specific library separately, for example, `library(dplyr)`.

</exercise>



<exercise id="2" title="Introduction to dplyr" type="slides">

<`r slides("chapter6_02_dplyr")`>

</exercise>

<exercise id="3" title="Check Your Understanding on `dplyr`">

**Problem 1**
In this problem, we will use the `iris` data set. The `iris` dataset is a built-in dataset in R that contains measurements on 4 different attributes (in centimeters) for 50 flowers from 3 different species: setosa, versicolor,virginica.

Create a new dataframe from the `iris` data set that meets the following criteria: contains only the `Sepal.Length` column and a new column called `petal_area` containing a value that approximates the area of the petal. (`Petal.Length` times `Petal.Width`). Only the rows where `petal_area` is greater than 5 should be shown in the final dataframe. The final dataframe should be called `new_iris`.

```{teachr 06_03_01}

new_iris <- iris %>% 
  <<mutate(petal_area = Petal.Length * Petal.Width) %>% filter(petal_area > 5) %>% select(Sepal.Length, petal_area)>>

---

Hint: You can use `mutate()`, `filter()` and `select()` to help with the task.
  
???

test <- function() {
    if (nrow(new_iris) != 117) {
        stop("Are you filtering the petal area correctly?")
    }
    success("Well done!")
}

test()

```

**Problem 2**

Joins
```{teachr 06_03_02}


---
  
Hint: 
  
???
  
```

</exercise>

<exercise id="4" title="Tidy (reshape) your data" type="slides">

<`r slides("chapter6_04_tidyr")`>

</exercise>

<exercise id="5" title="Check Your Understanding on `tidyr`">

**Problem 1**


Which table is tidy?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("Table 1 only" = "That's correct!",
         "Table 3 only" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` contains data with two values (one divided by the other) and is converted to characters.",
         "Table 1 and 3" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` in Table 3 contains data with two values (one divided by the other) and is converted to characters.",
         "Table 2 and 3" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` in Table 3 contains data with two values (one divided by the other) and is converted to characters. Table 2 contains two variables under the same column 'type'.",
         correct = 1))
```


**Problem 2**
Pivoting

```{teachr 06_05_01}

---
  
???

```
**Problem 3**

Missing values

```{teachr 06_05_02}

---
  
???

```


**Congratulations! This is the end of Data Wrangling!**
Here is a useful cheat sheet from the RStudio with commands that we introduced and many more:
https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf 
</exercise>





