---
title: 'Chapter 7: Linear regression with R'
description: |
  This chapter demonstrates how to fit linear models in R using the `lm` function. You'll learn about how to fitting simple linear regression, models with categorical variables and interaction effects, and visualising models with `ggplot2`.
prev: /chapter6
next: /chapter7
type: chapter
id: 7
output:
  teachr::teachr_chapter:
  html_document:
    self_contained: false
  #md_document
---

```{r setup, include = FALSE}
library(teachr)
knitr::opts_chunk$set(fig.path = "ch7/",
                      fig.retina = 2,
                      fig.width = 6, 
                      fig.height = 4,
                      fig.align = "center",
                      cache = TRUE,
                      cache.path = "cache/",
                      message = FALSE,
                      warning = FALSE)
```


<exercise id="1" title="Simple linear regression" type="slides">

`r slides("chapter7_01_simple-linear-regression")`

</exercise>

<exercise id="2" title="Practice extracting model estimates">


```{r}
library(Ecdat)
str(Cigarette)
```

This data was collected for each state from 1985 to 1995 in USA. The variables are described further below:

* `state`: state
* `year`: year
* `cpi`: consumer price index
* `pop`: state population
* `packpc`: number of cigarette packs per capita
* `income`: state personal income (total, nominal)
* `tax`:  average state, federal, and average local excise taxes for fiscal year
* `avgprs`: average cigarette price during fiscal year, including sales taxes
* `taxs`: average excise taxes for fiscal year, including sales taxes


A simple linear model for the average cigarette price on average excise taxes was proposed. 

Fit the model using least squares approach and calculate:

(a) residual sum of squares and assign it to `RSS`  
(b) least squares estimate of the intercept to `beta0`  
(c) least squares estimate of the slope to `beta1`  


```{teachr 07_02A, outdir = "../exercises"}
data(Cigarette, package = "Ecdat")

fit <- lm(<<avgprs ~ taxs>>, data = Cigarette)
  
RSS <- <<sum(residuals(fit)^2)>>
beta0 <- <<coef(fit)[1]>>
beta1 <- <<coef(fit)[2]>>

---

Hint: What did functions `coef` and `residuals` do?
  
???

test <- function() {
  fit <- lm(avgprs ~ taxs, data = Cigarette)
  expect_equal(RSS, sum(residuals(fit)^2))
  expect_equal(beta0, coef(fit)[1])
  expect_equal(beta1, coef(fit)[2])
}  
  
```



</exercise>



<exercise id="3" title="Regression with categorical variables" type="slides">

`r slides("chapter7_02_regression-with-categorical-var")`

</exercise>


<exercise id="4" title="Test your knoweldge">


</exercise>


<exercise id="5" title="Regression models with interaction effects" type="slides">
`r slides("chapter7_03_interaction")`
</exercise>


<exercise id="6" title="Practice">

</exercise>


<exercise id="7" title="Visualising regression models"  type="slides">
`r slides("chapter7_04_visualisation")`
</exercise>

<exercise id="8" title="Practice">

</exercise>
