---
title: 'Chapter 5: Data wrangling'
description:
  'In this chapter, you will learn to use tidyr and dplyr from the tidyverse in R.'
prev: /chapter4
next: /chapter6
type: chapter
id: 5
output:
  teachr::teachr_chapter:
  html_document:
    self_contained: false
    keep_md: true
---

```{r setup, include = FALSE}
library(teachr)
library(tidyverse)
knitr::knit_hooks$set(class = function(before, options, envir) {
  if(before) {
    sprintf("<div class='%s'>", options$class)
  } else {
    "</div>"
  }
})
knitr::opts_chunk$set(fig.retina = 2,
                      fig.width = 6, 
                      fig.height = 4,
                      fig.align = "center",
                      cache = TRUE,
                      cache.path = "cache/",
                      message = FALSE,
                      warning = FALSE)
```


<exercise id="1" title="Introduction to tidyverse">

## Introduction to tidyverse

The tidyverse is a collection of R packages that transform and visualize data. All packages of the tidyverse share very similar syntax and ideologies.

Tidyverse includes the following packages:

- `dplyr`: grammar of data manipulation
- `tidyr`: reshape your data
- `ggplot2`: grammar of graphics
- `readr`: read rectangular data
- `tibble`: re-imagining of data frame
- `stringr`: working with string data

and many more...


There are many other packages with more specialized functions in the tidyverse. In this chapter, we focuses on two packages: **tidyr** and **dplyr**.

**dplyr** is used for grammar of data manipulation. **tidyr** helps you to create tidy data or data where each variable is in a column, each observation is a row and each value is a cell. You can use them to solve the most common data manipulation challenges.

You can install the core tidyverse packages with:
`install.packages("tidyverse")`

Then, to load the package:
```{r}
library(tidyverse)
```

You can also load specific library separately, for example, `library(dplyr)`.

</exercise>



<exercise id="2" title="Introduction to dplyr" type="slides">

<`r slides("chapter5_02_dplyr")`>

</exercise>


<exercise id="3" title="Check Your Understanding on dplyr">

**Problem 1**

In this problem, we will use the `iris` data set. The `iris` dataset is a built-in dataset in R that contains measurements on 4 different attributes (in centimeters) for 50 flowers from 3 different species: setosa, versicolor,virginica.

Create a new dataframe from the `iris` data set that meets the following criteria: contains only the `Sepal.Length` column and a new column called `petal_area` containing a value that approximates the area of the petal. (`Petal.Length` times `Petal.Width`). Only the rows where `petal_area` is greater than 5 should be shown in the final dataframe. The final dataframe should be called `new_iris`.

```{teachr 05_03_01}

new_iris <- iris %>% 
  <<mutate(petal_area = Petal.Length * Petal.Width) %>% filter(petal_area > 5) %>% select(Sepal.Length, petal_area)>>

---

Hint: You can use `mutate()`, `filter()` and `select()` to help with the task.
  
???

test <- function() {
    if (nrow(new_iris) != 117) {
        stop("Are you filtering the petal area correctly?")
    }
    success("Well done!")
}

test()

```

**Problem 2**

Consider the data sets `size` and `color`. 

```{r}
size <- tibble(
  product_name = c('mini', 'pro', 'maxi'),
  diameter = c(23, 27, 32)
)

color <- tibble(
  product_name = c('mini','light', 'pro'),
  hue = c('red','green', 'silver')
)

```

We would like to merge the tables so that the result looks like:

```
# A tibble: 4 × 3
  product_name hue    diameter
  <chr>        <chr>     <dbl>
1 mini         red          23
2 light        green        NA
3 pro          silver       27
4 maxi         NA           32

```

Which of the following commands does the operation?

```{r, echo = FALSE,results = "asis"}
cat(mc_opts("<code>inner_join(color, size)</code>" = "Try again!",
        "<code>left_join(color, size)</code>" = "Try again!",
        "<code>right_join(color, size)</code>" = "Try again!",
        "<code>full_join(color, size)</code>" = "That's right!",
        correct = 4))
```

</exercise>

<exercise id="4" title="Tidy (reshape) your data" type="slides">

<`r slides("chapter5_04_tidyr")`>

</exercise>

<exercise id="5" title="Check Your Understanding on tidyr" >

**Problem 1**


We have used `table1`, `table2` and `table3` as example in the slides. Which table is tidy?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("Table 1 only" = "That's correct!",
         "Table 3 only" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` contains data with two values (one divided by the other) and is converted to characters.",
         "Table 1 and 3" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` in Table 3 contains data with two values (one divided by the other) and is converted to characters.",
         "Table 2 and 3" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` in Table 3 contains data with two values (one divided by the other) and is converted to characters. Table 2 contains two variables under the same column 'type'.",
         correct = 1))
```


**Problem 2**

Consider the following simple table `smoke`, which command do I need to transform `smoke` to `smoke2`?
```{r}
smoke <- tribble(
  ~smoke, ~employed, ~unemployed,
  "yes", 23, 10,
  "no", 20, 12
)
```

```
smoke2
# A tibble: 4 × 3
  smoke employment_status count
  <chr> <chr>             <dbl>
1 yes   employed             23
2 yes   unemployed           10
3 no    employed             20
4 no    unemployed           12
```

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("Using <code>pivot_wider()</code> on smoke" = "Try again!",
        "Using <code>pivot_longer()</code> on smoke" = "That's correct!",
        "Using <code>select()</code> on smoke" = "Try again!",
        "Using <code>filter()</code> on smoke" = "Try again!",
        correct = 2))
```

**Problem 3**

Which is the best way to check if there are missing observations in your data?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("Eyeballing and look for blanks in my data" = "That's not very effective nor efficient. Use R command to help you!",
        "Eyeballing and look for <code>NA</code> in my data" = "That's not very effective nor efficient. Use R command to help you!",
        "Using <code> is.na() </code> on my data" = "Try again!",
        "There is no need to find the missing values. We can ignore them." = "That's incorrect.",
        correct = 3))
```



**Congratulations! This is the end of Data Wrangling!**

Here is a useful cheat sheet from the RStudio with commands that we introduced and many more:
https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf 
</exercise>





