---
title: 'Chapter 6: Data Wrangling'
description:
  'In this chapter, you will learn to use tidyr and dplyr from the tidyverse in R.'
prev: /chapter5
next: /chapter7
type: chapter
id: 6
output:
  html_document:
    self_contained: false
    keep_md: true
---

```{r setup, include = FALSE}
library(teachr)
library(testthat)
```


<exercise id="1" title="Introduction to tidyverse">

## Introduction to tidyverse

The tidyverse is a collection of R packages that transform and visualize data. All packages of the tidyverse share very similar syntax and ideologies.

Tidyverse includes the following packages:

- `dplyr`: grammar of data manipulation
- `tidyr`: reshape your data
- `ggplot2`: grammar of graphics
- `readr`: read rectangular data
- `tibble`: re-imagining of data frame
- `stringr`: working with string data

and many more...


There are many other packages with more specialized functions in the tidyverse. In this chapter, we focuses on two packages: **tidyr** and **dplyr**.

**dplyr** is used for grammar of data manipulation. **tidyr** helps you to create tidy data or data where each variable is in a column, each observation is a row and each value is a cell. You can use them to solve the most common data manipulation challenges.

You can install the core tidyverse packages with:
`install.packages("tidyverse")`

Then, to load the package:
```{r}
library(tidyverse)
```

You can also load specific library separately, for example, `library(dplyr)`.

</exercise>



<exercise id="2" title="Introduction to dplyr">

`dplyr` provides a selection of tools for the most common data wrangling tasks.

## Data Transformation and Aggregation

There are a bunch of useful commands to transform and aggregate data in R.

- `select()`: pick particular variables from your data.

```{r}
table1 %>%
  select(country)
```

- `filter()`: extract observations from the data based on a given condition
```{r}
table2 %>%
  filter(year==1999)
```

- `arrange()`: sort your data based on one or more variables
```{r}
table2 %>%
  arrange(count)
```

- `mutate()`: add new variables or change existing variables
```{r}
table1 %>%
  mutate(rate = cases/population)
```

- `group_by()` and `ungroup()`: takes existing data and groups specific variables together for future operations. These two functions are often used together with other `dplyr` functions. We will introduce them with the `summarize()` function. The `summarize()` function creates a new dataframe based on the argument provided.

```{r}
table2 %>% group_by(type) %>% summarize(min = min(count), max =max(count))
```

You can summarize the data in different ways based on the groupings and different argument you included.
```{r}
table2 %>% group_by(type) %>% summarize(mean(count), median(count),sd(count))
```

- `count()`: a convenient way to get a sense of the distribution of values in a data set.
```{r}
table2 %>% count(country)
```

## Joining Datasets

Often you will need information from multiple datasets in your data analysis. It is important to learn different types of joins to accommodate different needs in your tasks.

We are going to use the following tables in our example. Consider there are 4 students in a class and they submitted homework 1 and homework 2. The tables contains their scores for the homeworks (gradebook) and not everyone submitted their homework during the semester.

```{r}
homework1 <- tibble(name = c('Alphonso', 'Daisy', 'Philip'), student_no = c(456,100,102), marks = c(80, 65, 34))
homework2 <- tibble(name = c('Alphonso', 'Philip', 'Melinda'), student_no = c(456,102,432), marks = c(85, 97, 56))
```

### Mutating Joins

Mutation joins return a copy of the dataset that has been augmented.

**Inner Join**

Inner joins retain only rows that exists on both datasets.
```{r}
inner_join(homework1, homework2, by = c('name','student_no'))
```
When we perform inner join by `name` and `student_no`, R only returns rows that exists in both data set. When we do inner join with our `homework1` and `homework2` gradebooks, only `Alphonso` and `Philip` 's marks will show up since they are the only ones that submitted both homeworks.

**Left Join**

Left joins starts with the left data set (`homework1`) and search in the right data set looking for the same values of `name` and `student_no` to add to the left data set. 
```{r}
left_join(homework1, homework2, by = c('name','student_no'))
```
When we do left join with our `homework1` and `homework2` gradebooks, we will only consider those students who submitted homework 1 and look for their marks in homework 2. Melinda's result will not be included in the join even though she has submitted her homework 2, while Daisy will have `NA` showing as the missing value.

**Right Join**

Right joins starts with the right data set (homework2) and search in the left data set looking for the same values of `name` and `student_no` to add to the right data set.
```{r}
right_join(homework1, homework2, by = c('name','student_no'))
```
When we do right join with our `homework1` and `homework2` gradebooks, we will only consider those students who submitted homework 2 and look for their marks in homework 1. Daisy's result will not be included in the join even though she has submitted her homework 1, while Melinda will have `NA` showing as the missing value.


**Full Join**

Full joins retain all rows.
```{r}
full_join(homework1, homework2, by = c('name','student_no'))
```
All the submissions with the corresponding mark are included in the full join of the gradebook.

### Filtering Joins

Filtering joins return a copy of the dataset that has been filtered.

**Semi Join**

The semi join returns a filtered version of the left data set based on matching entries in the right data set.
```{r}
semi_join(homework1, homework2, by = c('name','student_no'))
semi_join(homework2, homework1, by = c('name','student_no'))
```

**Anti Join**

Now, let's say we want to know who didn't submit homeworks. The anti join will return the rows in the left data set that doesn't have an entry in the right data set.
```{r}
anti_join(homework1, homework2, by = c('name','student_no'))
anti_join(homework2, homework1, by = c('name','student_no'))
```

</exercise>

<exercise id="3" title="Check Your Understanding on `dplyr`">

</exercise>

<exercise id="4" title="Tidy (reshape) your data">

## Tidy (reshape) your data

Raw data that we acquired can often be messy and complicated. The tidyr package allows you to wrangle messy datasets into nice and tidy ones.

A tidy data format has a rectangular shape which means it has columns, rows, and cells, just like in a spreadsheet. Each column should hold a single variable, each row should hold a single observation and each cell should hold a single value.

The example below shows the same information (country, year, population, and number of cases) organised in 3 different ways.
Examine each of the table.

```{r}
table1
table2
table3
```

Which table is tidy?

```{r, echo = FALSE, results = "asis"}
cat(mc_opts("Table 1 only" = "That's correct!",
         "Table 3 only" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` contains data with two values (one divided by the other) and is converted to characters.",
         "Table 1 and 3" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` in Table 3 contains data with two values (one divided by the other) and is converted to characters.",
         "Table 2 and 3" = "Try again! Each cell in a tidy data format should hold only a single value. The column `rate` in Table 3 contains data with two values (one divided by the other) and is converted to characters. Table 2 contains two variables under the same column 'type'.",
         correct = 1))
```

Sometimes raw data may have multiple variables or multiple values (not the same variable) in the same column, or sometimes you may want to merge multiple columns into one. Here are a few commands that can be helpful in manupulating the data into a tidy format:

- `separate()`: separating multiple variables in one column to different columns
- `unite()`: merging multiple columns into one, inserting underscore `_` as a separator between the two inputs.
- `separate_rows()`: separating multiple observations in the same cell. We pass the messy column name and a string to use as a separator.

In table 3, the `rate` column contains both the number of cases and the size of the populations.
We want to separate the `rate` variable into variables `cases` and `population`. 

```{r}

table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)

```
The `into` argument expects a vector of column names. The `sep` argument expects the non-alphanumeric character (a charater that is not a letter or a number) that we split upon. The `convert` argument is set to TRUE here so that R will try and convert to better types after the split.


In table 5, the `century` and `year` column can be combined to give a `year_new` column.
The `sep` argument here is what comes in between the values from different columns. If we don't want anything in between, we can use "".

```{r}

table5 %>%
  unite(new_year,century, year, sep = "")

```

In table 6, there are multiple observations in the same cell in each row of `year` and `cases`. We would like to separate them.
```{r}

table6 <- tibble(
  country = c("Afghanistan", "Brazil", "China"),
  year = c("1999,2000", "1999,2000", "1999,2000"),
  cases= c("745,2666", "37737,80488", "212258,213766")
)

table6 %>%
  separate_rows(cases, year, sep=',', convert = TRUE)
```

## Pivoting

Two common problems in raw data set:
- some of the column names are not really names of variables, but **values** of a variable. 
- when an observation is scattered across multiple rows.

To tackle these problems, you will need `pivot_longer()` and/or `pivot_wider()`

**pivot_longer()**

`table4a` is a great example of data that needs `pivot_longer()`. Here you see that the case count of each country is separated into two columns because they are from different years.
```{r}
table4a
```

The first argument is the vector of column names that are actually values not variables.
The second argument is the name of the variable to move the column **names** to.
The third argument is the name of the variable to move the column **values** to.

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

**pivot_wider()**

In table 2, each observation is a certain country in a certain year. However, the observation is spread across two rows since there are `cases` and `population`.

```{r}
table2
```

The first argument is the column to take variable **names** from, which is `type`.
The second argument is the column to take variable **values** from, where is `count`.

```{r}
table2 %>%
  pivot_wider(names_from = type, values_from = count)
```


## Missing values

There are various ways to handle missing values (the NAs) in your data set. Tidyr has three functions:

- `fill()`: fill the missing values using the last observation carried forward. It has a `.direction` argument that you can pass `up` or `down`.
- `drop_na()`: drop the rows with observations that are `NA`
- `replace_na()`: replace the `NA` with specific values

Let's look at this toy data set. We will start with the `fill()` function to impute the `return` column in the `up` direction, then we will impute in the `down` direction. Can you spot the difference?

```{r}
apple <- tibble(
  year   = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(0.75, 0.39, 0.34,   NA, 0.99, 0.07, 1.88)
)

apple %>%
  fill(return, .direction = "up")

apple %>%
  fill(return, .direction = "down")
```


Then, we will try to drop the NAs in the return in `apple`.

```{r}
apple %>%
  drop_na(return)
```
 
Finally, we will also replace the NAs with the value `999`.
```{r}
apple %>%
  replace_na(list(return=999))
```


</exercise>

<exercise id="5" title="Check Your Understanding on `tidyr`">

</exercise>





